---
title: "Proteomics analysis pipeline results"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(drake)
library(DT)
library(kableExtra)
library(knitr)
```
  
```{r load targets}
loadd(data_in)
loadd(wilcox_results)
loadd(fold_change_results)
loadd(plot_results)
loadd(random_forest_results)
loadd(proteins_ranked)
loadd(unique_protein_list)
loadd(GO_term_analysis_results)
loadd(pathway_analysis_results)
loadd(disease_analysis_results)
```

## Data

```{r data info display}
names <- c("Number of features", "Number of samples", "Class 1", "Class 2",
  "Number of samples in class 1", "Number of samples in class 2")

values <- as.character(c(dim(data_in)[2] - 2, dim(data_in)[1], levels(data_in[, 2])[1], 
  levels(data_in[, 2])[2],
  nrow(data_in[which(data_in[, 2] == levels(data_in[, 2])[1]), ]),
  nrow(data_in[which(data_in[, 2] == levels(data_in[, 2])[2]), ])))

info <- cbind(names, values)

DT::datatable(info, options = list(dom = 't', ordering = FALSE), colnames = rep("", ncol(info)), )
```

## First analysis {.tabset}

### Wilcoxon rank-sum test

```{r wilcoxon results NA check}

#check for NA values in data frame
wilcoxCounterNA <- sum(is.na(wilcox_results[, 4]))
  
#number of proteins for which p-value was calculated
N <- nrow(wilcox_results) - wilcoxCounterNA

#calculate number of significant proteins
wilcoxResultsSignificant <- wilcox_results %>% 
  filter(Pvalue.adj < 0.05) %>% 
  arrange(Pvalue.adj)

M <- nrow(wilcoxResultsSignificant)

#`r if(wilcoxWarningNA){paste("WARNING:", NAcounter, "NA values detected.", sep = " "}`
```

The wilcoxon test failed for `r wilcoxCounterNA` proteins. 

P-value calculated for **`r N`** proteins. Out of those **`r M`** proteins showed a significant p-value < 0.05. The complete list of significant proteins can be found below.

```{r wilcoxon results table DT}
DT::datatable(wilcoxResultsSignificant, colnames = c("Description", "Gene", "P-value", 
  "Adjusted P-value")) %>% 
  formatRound(c("Pvalue", "Pvalue.adj"), 6)
```

### Fold change 

```{r fold change results NA check}
#check for NA values in data frame
foldChangeCounterNA <- sum(is.na(fold_change_results[, 2]))
```

The fold change calculation failed for `r foldChangeCounterNA` proteins. 

The complete list of the calculated fold change between affected and control samples can be found below.

```{r fold change results table}
# fold_change_results  %>%
# kable(format = "html") %>%
# kable_styling()
DT::datatable(fold_change_results, colnames = c("Description", "Gene", "Fold change")) %>% 
  formatRound("foldChange", 6)
```

### Volcano plot
```{r volcano plot}
plot_results
```

```{r volcano plot marked proteins, eval = FALSE}
indSignificant <- which((results$Pvalue.adj < 0.05 & results$foldChange > 0.25) |
    (results$Pvalue.adj < 0.05 & results$foldChange < -0.25))
resultsSignificant <- results[indSignificant, ]
```

### Info

## Random Forest Classification {.tabset}

### Model accuracy

```{r random forest accuracy}
testError <- vector()

for (i in 1:length(random_forest_results)) {
    #subset one random forest object
    fit.RF <- random_forest_results[[i]]
    #extract feature importance ranking and append to results vector
    testError <- append(testError, fit.RF$test$err.rate[fit.RF$ntree])
}

accuracy <- sapply(testError, FUN = function(x) (1 - x)*100)
```

The classification models produced an average accuracy of **`r round(mean(accuracy), 2)`** with a standard deviation of **`r round(sd(accuracy), 2)`**. The distribution of the individual model accuracies is shown in the following boxplot.

```{r random forest accuracy plot}
ggplot(as.data.frame(accuracy), aes(x="", y = accuracy)) + 
  geom_boxplot(colour = "black", fill = "dodgerblue4") +
  labs(x = "", y = "accuracy", title = "Classification accuracy of Random Forest models")
```

### Protein ranking

```{r proteins ranked table}
# proteins_ranked  %>%
# kable(format = "html") %>%
# kable_styling()

DT::datatable(proteins_ranked, rownames = FALSE) %>% 
  formatRound("Importance", 6)
```


### Protein set selection

`r nrow(unique_protein_list)` were selected for further analysis.

```{r proteins selected importance table}
# unique_protein_list  %>%
# kable(format = "html") %>%
# kable_styling()

DT::datatable(unique_protein_list, rownames = FALSE)
```

### Highest ranking proteins

```{r proteins selected importance plot}
ranking_plot <- proteins_ranked[1:10, 2:3]

ranking_plot$Gene <- factor(ranking_plot$Gene,
                        levels=ranking_plot$Gene)

ggplot(ranking_plot, aes(x = ranking_plot[, 1], y = ranking_plot[, 2])) +
  geom_bar(aes(), stat="identity", colour = "black", fill = "dodgerblue4") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(x = "Genes", y = "Importance", title = 
      "Genes with highest average importance across all Random Forests")
```
### Info 

## Protein set analysis {.tabset}

### Gene ontology term analysis
  
```{r GO term analysis}
# as.data.frame(GO_term_analysis_results) %>%
#   kable(format = "html") %>%
#   kable_styling()

DT::datatable(as.data.frame(GO_term_analysis_results), rownames = FALSE, 
  colnames = c("ID", "Description",   "Gene Ratio", "BG Ratio", "P-value", "Adjusted P-value", 
  "Q-value", "Gene ID")) %>% 
  formatRound(c("pvalue", "p.adjust", "qvalue"), 6)
```


### Pathway enrichment analysis
  
```{r pathway enrichment analysis}
# data.frame(pathway_analysis_results) %>%
#   kable(format = "html") %>%
#   kable_styling()

DT::datatable(data.frame(pathway_analysis_results), rownames = FALSE, 
  colnames = c("ID", "Description", "Gene Ratio", "BG Ratio", "P-value", "Adjusted P-value", 
  "Q-value", "Gene ID")) %>% 
  formatRound(c("pvalue", "p.adjust", "qvalue"), 6)
```


### Disease ontology analysis 
  
```{r disease analysis}
# data.frame(disease_analysis_results) %>%
#   kable(format = "html") %>%
#   kable_styling()

DT::datatable(data.frame(disease_analysis_results), rownames = FALSE, 
  colnames = c("ID", "Description", "Gene Ratio", "BG Ratio", "P-value", "Adjusted P-value", 
  "Q-value", "Gene ID")) %>% 
  formatRound(c("pvalue", "p.adjust", "qvalue"), 6)
```

### Info
